---
title: "Case study: How does a bike-share navigate speedy success?"
author: "Nisha P"
date: "2024-03-31"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Cyclistic, a prominent bike-share company headquartered in Chicago, has
rapidly gained traction in the city's transportation landscape. In an
effort to delve deeper into their customer base and refine their
marketing strategies, Cyclistic seeks to understand the distinct
behaviors and preferences of casual riders versus annual members.

The company recognizes the need to tailor its marketing approach to
effectively convert casual riders into committed annual members. By
leveraging data-driven insights, Cyclistic aims to develop a
comprehensive understanding of how these two customer segments interact
with their services differently.

**Business Task**

The objective of this business task is to develop a comprehensive
marketing strategy for Cyclistic that addresses the distinct needs and
behaviors of both annual members and casual riders. By answering the
following three questions, we aim to optimize marketing efforts,
increase customer engagement, and drive conversions from casual riders
to annual members.

-   ***Understanding Usage Patterns***: Analyze Cyclistic's dataset to
    identify differences in how annual members and casual riders utilize
    Cyclistic bikes.

**Data Background**

The dataset was acquired from [Click
here](https://divvytripdata.s3.amazonaws.com/index.html) and *Motivate
International Inc* made the data available under this
[license](https://divvybikes.com/data-license-agreement).

For this project, I downloaded data for twelve months (January to
December 2020). The zipped CSVs were downloaded and unzipped into a
folder.

Below shown the dataset of a *cyclistic biketrip data for the year
2020*.The dataset has *3541683* rows and *13* column.

Due to the large size of data we use *R* to analyse effectively.

## R Programming

-   ***Loading Packages*** The R package is a collection of R functions,
    data sets, and compiled code that extends the functionality of R.
    Here we use four packages to analyse the data.

In R, the *library()* function is used to load R packages into your
current R session

```{r}
library (tidyverse)
library (janitor)
library (lubridate)
library (scales)
library(readr)
library(geosphere)
rm(list=ls())
```

-   **Read CSV file**

Below given are the year ***2020 dataset** of cyclistic bike share
program* which are downloaded and saved as CSV files. Here
***read.csv()*** is used for reading the csv files.

df1 \<- read.csv("Divvy_Trips_2020_Q1.csv") df2 \<-
read.csv("202004-divvy-tripdata.csv") df3 \<-
read.csv("202005-divvy-tripdata.csv") df4 \<-
read.csv("202006-divvy-tripdata.csv") df5\<-
read.csv("202007-divvy-tripdata.csv") df6 \<-
read.csv("202008-divvy-tripdata.csv") df7 \<-
read.csv("202009-divvy-tripdata.csv") df8 \<-
read.csv("202010-divvy-tripdata.csv") df9 \<-
read.csv("202011-divvy-tripdata.csv") df10 \<-
read.csv("202012-divvy-tripdata.csv") df20 \<-
rbind(df1,df2,df3,df4,df5,df6,df7,df8,df9,df10)

Save the binded dataset as CSV file.

write.csv(df20,file = "df20.CSV",row.names = FALSE)

```{r}
df20 <- read_csv("C:/Users/nisha/Desktop/New folder/DataAnalytics_NishaP/Dataset/df20.CSV")
```

In R, the ***head()*** function is used to view the first few rows of a
data frame or a matrix. It allows you to quickly inspect the structure
and content of your data without displaying the entire dataset.

```{r}
head(df20)
```

-   **Cleaning Data**

***Janitor*** is an R package that provides a set of functions to clean
and preprocess data in R data frames

```{r}
df20_cleanedcols <- janitor::remove_empty(df20,which =c("cols"))
df20_cleanedrows <- janitor::remove_empty(df20,which =c("rows"))
dim(df20_cleanedcols)
dim(df20_cleanedrows)
```

-   **Removing duplicates and NA values**

```{r}
df20_clean <- na.omit(df20)
# for unique and removing duplicates 
unique(df20_clean)
dim(df20_clean)
df20_clean <- df20_clean %>% filter(df20_clean$start_station_name!=" ")
```

-   **Organising Data**

*Lubridate* is an R package designed to make it easier to work with
dates and times in R. It provides a set of functions that simplify
common tasks such as parsing, manipulating, and formatting dates and
times.we use parse date *ymd_hms()* and *as.Date()* for changing the
Started_at and ended_at column format.

***Difftime()*** is used for calculating the difference in time. This
helps us to find and analyse the duration of each ride.

```{r}
df <- df20_clean

df$started_date <- as.Date(df$started_at)
df$ended_date <- as.Date(df$ended_at)
#time as  hours and minutes
df$started_at <- lubridate::ymd_hms(df$started_at)
df$ended_at <- lubridate::ymd_hms(df$ended_at)
df$Start_time <- format(df$started_at,"%H:%M:%S")
df$End_time <- format(df$ended_at,"%H:%M:%S")
df$day_of_the_week <- weekdays(df$started_at)
df$month <- month(df$started_at, label = TRUE, abbr = TRUE)

df$trip_duration <- (as.double(difftime(df$ended_at,df$started_at)))/60

df<-df %>% 
  filter(trip_duration > 0)

glimpse(df)
```

Dim function ***dim(df)*** retrieve or set the dimensions of an object,
such as a matrix or an array.

Here's how it works:

```{r}
dim(df)
```

The distHaversine functionin R, from the geosphere package, is used to
calculate the great-circle distance between two points on the Earth's
surface given their latitude and longitude coordinates. This distance is
calculated using the Haversine formula, which accounts for the spherical
shape of the Earth.

```{r}
df$distance <- mapply(function(lat1, lon1, lat2, lon2) {
  distHaversine(c(lon1, lat1), c(lon2, lat2))
}, df$start_lat, df$start_lng, df$end_lat, df$end_lng)

#change to km
df$distance <- df$distance/1000

glimpse(df)


```

-   **Summarize the data**

## summary of the dataframe by removing unnecessary column

```{r}
df <- df %>% 
  select(-start_station_id,-end_station_id,-start_lat,-end_lat,-start_lng,-end_lng)
glimpse(df)
```

```{r}
##calculate riable_type usage
sum_df <- df %>% 
  select(rideable_type,member_casual,started_at,start_station_name,day_of_the_week,month,trip_duration,distance) %>% 
  group_by(rideable_type,member_casual) %>% 
  summarise(Total_Duration = sum(trip_duration),Count = n(),Total_distance = sum(distance)) %>% 
  ungroup()

glimpse(sum_df)
```

```{r}
##calculate rideable_type usage
sum_df <- df %>% 
  select(rideable_type,member_casual,started_at,start_station_name,day_of_the_week,month,trip_duration,distance) %>% 
  group_by(rideable_type,member_casual) %>% 
  summarise(Total_Duration = sum(trip_duration),Count = n(),Total_distance = sum(distance)) %>% 
  ungroup()

glimpse(sum_df)
```

```{r}
## member Vs Casual distribution
Member_type<- df %>% 
  group_by(member_casual) %>% 
  summarise(Count = n(),Total_duration = sum(trip_duration),Total_distance = sum(distance)) %>% 
  ungroup()
glimpse(Member_type)
```

```{r}
# Daily ride 
ride_per_day <- df %>% 
  group_by(started_date,member_casual) %>% 
  summarise(Avg_Trip = mean(trip_duration),Avg_distance = mean(distance),Count = n()) %>% 
  arrange(started_date) %>% 
  ungroup()


glimpse(ride_per_day)

```

```{r}
## weekly ride
Weekly_ride <- df %>% 
  group_by(day_of_the_week,member_casual) %>% 
  summarise(Avg_Trip = mean(trip_duration),
            Avg_distance = mean(distance),Count = n()) %>% 
  arrange(day_of_the_week) %>% 
  ungroup()

glimpse(Weekly_ride)
```

```{r}
# monthly Ride
monthly_ride <- df %>% 
  group_by(month, member_casual) %>% 
  summarise(Avg_Trip = mean(trip_duration),
            Avg_distance = mean(distance),Count = n()) %>% 
  arrange(month) %>% 
  ungroup()


glimpse(monthly_ride)
```

```{r}
# Popular start station 
Popular_top_start_stations <- df %>%
  count(start_station_name) %>%
  arrange(desc(n)) %>% 
  head(10)
```

```{r}
# Top 20 start station
top_start_stations <- df %>%
  group_by(start_station_name,member_casual) %>% 
  count(start_station_name) %>%
  arrange(desc(n)) %>% 
  head(20) %>% 
  ungroup()
```

```{r}
#Top 20 end Station
  top_end_stations <- df %>%
  group_by(end_station_name,member_casual) %>% 
  count(end_station_name) %>%
  arrange(desc(n)) %>% 
  head(20) %>% 
 ungroup()
```

```{r}
head(top_start_stations)
head(top_end_stations)
```

```{r}
#  top station with large distance ride 
dis_df <- df %>% 
  group_by(start_station_name,member_casual) %>% 
  summarise(Avg_distance = mean(distance)) %>% 
  arrange(desc(Avg_distance)) %>% 
  head(20) %>% 
  ungroup()

head(dis_df)

```

```{r}
#hourly Bike Demand

df <- df %>% 
  mutate(start_hour = lubridate::hour(started_at))
```

```{r}
hourly_need <- df %>% 
 group_by(member_casual,start_hour) %>% 
  summarise(number_of_trips = n()) %>% 
  ungroup()
```

-   **Data Visualisation**

The *ggplot()* function is the primary function used in the ggplot2
package, a popular data visualization package in R. It is used to create
and customize plots based on a grammar of graphics approach, allowing
users to create complex and highly customizable visualizations with
relatively simple syntax.

Here's how the ggplot() workes to calculate

1.  ***Member Vs Casual Distribution***

```{r}
Member_type$Percentage <- round(Member_type$Count/sum(Member_type$Count)*100)

ggplot(Member_type,mapping = aes(x = " ", y = Percentage, fill = member_casual)) + 
    geom_col(color = "black") + 
    geom_text(aes(label=paste(member_casual, paste(Percentage,"%"),sep="\n")),     position = position_stack(vjust=0.5), color="black") +
  labs(title = "Members vs Casual Distribution") +
  coord_polar(theta = "y") +
  scale_fill_brewer() +
  theme_bw()
```

2.  ***Riders Bike type Usage***

```{r}
#Most used bike type 
 ggplot(sum_df,mapping = aes(x = rideable_type ,y = Count,fill = rideable_type)) +
  geom_bar(stat = "identity") +
  facet_wrap(~member_casual, nrow = 1) +
  theme(legend.position = "none") +
  labs(title = "Rider Bike type Usage",x = "Bike_type",y = "Count")
```

3.  ***Rides Done on Per Day***

```{r}
# Total Ride per day

ggplot(ride_per_day, aes(x = started_date, y = Count,fill = factor(member_casual))) +
  geom_col()+ labs( title  ="Ride taken Per Day",
       x = "Date",
       y = "Count") +
  theme_minimal()
```

4.  ***Weekly Ride Count***

```{r}
#Total Ride per Week

ggplot(Weekly_ride, aes(x = day_of_the_week, y = Count,fill = factor(member_casual))) +
  geom_col()+ labs( title  ="Weekly Ride Count",
       x = "Day of Week",
       y = "Count") +
  theme_minimal()
```

5.  ***Monthly Ride Count***

```{r}

ggplot(monthly_ride, aes(x = month, y = Count ,fill = factor(member_casual))) +
  geom_col()+ labs( title  ="Monthly Ride Count",
       x = "Month(year 2020)",
       y = "Count") +theme_minimal()+
  theme(axis.text.x = element_text(angle = 45))
```

6.  **Average Distance Covered from Various Start Station**

```{r}
## Average Distance ride by member type in a year

ggplot(dis_df, aes(x = Avg_distance, y = reorder(start_station_name,Avg_distance), fill = factor(member_casual))) +
  geom_col() +
  labs(title = "Large Distance Covered from Various Start Station(Top 20)",
       x = "Distance Covered (km)",
       y = "Station Name",
       fill = "Rider Type") +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90))
```

7.  **Popular Start Station**

```{r}
Popular_top_start_stations %>% 
ggplot() + geom_col(aes(x=n,y= reorder(start_station_name,n)))+ scale_x_continuous(labels = comma)+
  labs(title = "Top 10 popular Start Station", y = "No of Rides")
```

8.  **Rider most preferred Stations**

```{r}
#Top Start Station Name

ggplot(top_start_stations, aes(x = n, y = reorder(start_station_name,n), fill = factor(member_casual))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 20 Start Station Name",
       x = "Ride Count",
       y = "Station Name",
       fill = "Rider Type") +
  theme_minimal()
```

```{r}
#Top End Station Name

ggplot(top_end_stations, aes(x = n, y = reorder(end_station_name,n), fill = factor(member_casual))) + 
  geom_bar(stat = "identity", position = "dodge") + 
  labs(title = "Top 20 End Station Name",
       x = "Ride Count",
       y = "Station Name",
       fill = "Rider Type") +
  theme_minimal()
```

9.  **Hourly Bike Demand**

```{r}
hourly_need %>% 
  ggplot()+geom_line(aes(x=start_hour,y= number_of_trips,color = member_casual))+
  labs(title = "Hourly Bike Demand",
       x = "Hour",
       y = "No of Trips",
       fill = "Member_casual") +
  scale_x_continuous(limits = c(0,24), name ="Hours")+
  theme_minimal()
```

## Recommendations for Stakeholders:

1.***Enhance Membership Programs:***

Since 62%\* of the users are members, there is an opportunity to further
strengthen membership benefits to retain and attract more long-term
users. Consider offering loyalty programs, discounts for long-term
memberships, or exclusive benefits during peak seasons (April to
September).

2.***Bike Type Optimization:***

Given that both members and casual riders prefer docked bikes, ensure
that there are sufficient docked bikes available at high-demand
stations, especially during peak hours. Consider investing in more
docked bikes and maintaining a balance with other bike types.

3.***Seasonal Promotions:***

Since ride frequency increases from April and decreases after September,
plan for seasonal promotions and marketing campaigns to maximize
ridership during these months. This could include discounted rides,
special events, or partnerships with local businesses to encourage more
usage.

4.***Improve Weekend Services:***

With higher ride volumes on weekends, ensure there are adequate
resources and bike availability. Consider running special weekend events
or promotions to further boost ridership.

5.***Focus on High-Demand Stations:***

Vincennes Ave & 104th Station and Streeter Dr & Grand Ave are key
stations with high ride counts and distances covered. Enhance services
at these stations, such as better bike maintenance, increased docking
stations, and potentially setting up customer service points.

6.***Adjust for Peak Hours:***

With peak demand between 3 pm to 6 pm, allocate more bikes and ensure
efficient redistribution of bikes to meet demand. Consider offering
incentives for riders who choose to ride outside of these peak hours to
balance the load.

### Recommendations for Marketing Team:

1.  ***Targeted Marketing Campaigns:***

Develop targeted campaigns to convert casual riders to members.
Highlight the benefits of membership, such as cost savings, exclusive
access to promotions, and convenience.

2.  ***Leverage Popular Stations:***

Use the popularity of stations like Streeter Dr & Grand Ave to create
event-based marketing. For instance, set up pop-up events, offer free
refreshments, or partner with nearby attractions to draw in more riders.

3.  ***Promote During Peak Seasons:***

Utilize data showing increased rides from April to September to launch
time-limited offers and campaigns. Engage with riders through social
media, email newsletters, and local advertisements to promote these
offers.

4.***Weekend Specials:***

Since weekends see higher ridership, promote special weekend passes or
family packages to attract group rides. Collaborate with local tourist
attractions or restaurants to offer combined deals.

5.***Highlight Environmental Impact:***

Emphasize the environmental benefits of using bike share programs in
your marketing materials. Share statistics on carbon footprint reduction
and promote the sustainable aspect of biking to attract eco-conscious
riders.

6.***Dynamic Pricing:***

Implement dynamic pricing strategies during peak hours and seasons to
manage demand and encourage off-peak usage. Offer discounted rates for
rides starting early in the morning or late at night.

By addressing these, Cyclistic can enhance user experience, optimize
operations, and effectively increase ridership and membership.

Thank you,

Nisha Prasanth.
